Peer Assignment 1 - Reproducible Research
========================================================

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

1. steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. date: The date on which the measurement was taken in YYYY-MM-DD format
3. interval: Identifier for the 5-minute interval in which measurement was taken

### Loading & processing data

*Below steps are performed by the r code chunk given below*

1. Read the CSV file into R using the following code, assumes the activity csv file is present in the working directory.

2. Transform the column/variable 'date' into date format, using the below format

3. Use tapply to get the sum of steps for each day

```{r}
        library(ggplot2)
        activity <- read.csv("activity.csv", stringsAsFactors=FALSE)

        activity$date <- as.Date(activity$date, "%m/%d/%Y")

        stepsSUM <- tapply(activity$steps, activity$date, sum)

        sumDF <- data.frame(day = names(stepsSUM), total = stepsSUM)        

```

## What is mean total number of steps taken per day?

*Below steps are performed by the r code chunk given below*

1. Make a histogram of the total number of steps taken each day

2. Calculate and report the mean and median total number of steps taken per day



```{r}
        ggplot(data = sumDF,aes(x=day,y=total)) + 
        geom_bar(stat="identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1,size=10,face="bold",vjust=1.0)) + 
        labs(x= "Day") + labs(y="Total Steps") + 
        labs(title =" Histogram Of The Total Number Of Steps Taken Each Day") 
        meanSteps       <-      mean(sumDF$total, na.rm=TRUE)
        meanSteps       <-      format(round(meanSteps,2), nsmall=2)
        meadianSteps    <-      median(sumDF$total, na.rm=TRUE)

```
*Mean of total number of steps taken    = `r meanSteps`.*
*Median of total number of steps taken  = `r meadianSteps`.*


## What is the average daily activity pattern?

*Below steps are performed by the r code chunk given below*

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
        avgSteps <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
        avgStepsDF <- data.frame(interval = names(avgSteps), avg = avgSteps)
        plot(avgStepsDF$interval, avgStepsDF$avg, xlab="5-min Interval", ylab="Avg Of Steps")

        maxInterval <- which.max(avgStepsDF$avg)
        
```

**The 5-minute interval that contains maximum number of steps is `r avgStepsDF$interval[maxInterval]`.**

## Imputing missing values

*Below steps are performed by the r code chunk given below*

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

2. Missing values are filled by the mean of that interval

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
        missingValues <- length(which(is.na(activity)))
        indexNA <- which(is.na(activity))
        IMactivity <- activity
        IMactivity$steps[indexNA] <- avgStepsDF$avg[avgStepsDF$interval[1:nrow(avgStepsDF)] %in%   IMactivity$interval[indexNA]]
        stepsBYdate <- tapply(IMactivity$steps, IMactivity$date, sum)
        stepsBYdateDF <- data.frame(day = names(stepsBYdate), total = stepsBYdate)
                
        ggplot(data = stepsBYdateDF,aes(x=day,y=total)) +
        geom_bar(stat="identity") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1,size=10,face="bold",vjust=1.0)) +
        labs(x= "Day") +
        labs(y="Total Steps") +
        labs(title =" Histogram Of The Total Number Of Steps Taken Each Day") 
        
        

        
        mean(stepsBYdateDF$total)
        median(stepsBYdateDF$total)

```
**Total number of missing values in the dataset == `r missingValues`**

## Impact of imputing

1. Mean of the dataset has not changed before and after computing (10766.19) 

2. Median of the dataset before imuting was 10765 and after imputing it is 10766.19 

## Are there differences in activity patterns between weekdays and weekends?

*Below steps are performed by the r code chunk given below*

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
        IMactivity$dayType <- ifelse(weekdays(IMactivity$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
        IMactivity$dayType <- as.factor(IMactivity$dayType)
        IMactivityWD <- subset(IMactivity, IMactivity$dayType == "weekday")
        IMactivityWE <- subset(IMactivity, IMactivity$dayType == "weekend")
        
        avgStepsWD <- tapply(IMactivityWD$steps, IMactivityWD$interval, mean)
        avgStepsWDDF <- data.frame(interval = names(avgStepsWD), avg = avgStepsWD)
        avgStepsWDDF$dayType[1:nrow(avgStepsWDDF)] <- "weekday"

        avgStepsWE <- tapply(IMactivityWE$steps, IMactivityWE$interval, mean)
        avgStepsWEDF <- data.frame(interval = names(avgStepsWD), avg = avgStepsWD)
        avgStepsWEDF$dayType[1:nrow(avgStepsWEDF)] <- "weekend"

        avgStepsIMDF <- rbind(avgStepsWEDF, avgStepsWDDF)
        avgStepsIMDF$dayType <- as.factor(avgStepsIMDF$dayType)

        library(lattice)
        xyplot(avg ~ interval | dayType, data = avgStepsIMDF)


```


### Based on the plot there is no differences in activity patterns between weekdays and weekends